# Tutorial 2: Access single-cell datasets from `scRNAseq` collection and annotate cell types

Welcome to this tutorial on accessing public single-cell datasets using `scRNAseq` and language-agnostic representations from `ArtifactDB`. We will showcase how to integrate and process single-cell datasets across languages, such as R and Python, and how to annotate cell types using reference datasets. Let's dive into the process!


## Introduction

The [scRNAseq](https://bioconductor.org/packages/devel/data/experiment/html/scRNAseq.html) package provides access to public single-cell RNA-seq datasets for use by other Bioconductor packages and workflows. This package has been updated to store datasets in language-agnostic representations described in [ArtifactDB](https://github.com/artifactdb), enabling easy access to datasets and analysis results across multiple programming languages such as R and Python.

The Python equivalent of the package is available [here](https://github.com/BiocPy/scrnaseq).

## Installation

Install the package from Bioconductor for R or from PyPI for Python:

::: {.panel-tabset}

## R
```r
BiocManager::install("scRNAseq", repos='http://cran.us.r-project.org')
```

## Python (shell)
```sh
pip install scrnaseq
```

:::

## Find Datasets

Each dataset is decorated with metadata such as the study title, species, number of cells, etc., to facilitate discovery. Let's see how we can list and search for datasets.

<a name="list-all-datasets"></a>
### List All Datasets

The `list_datasets()` function in Python or `surveyDatasets()` in R will display all available datasets along with their metadata.

::: {.panel-tabset}

## R
```{r}
suppressWarnings(library(scRNAseq))
all_ds <- surveyDatasets()
head(all_ds[, c("name", "title", "version")], 3)
```

## Python
```{python}
import scrnaseq
datasets = scrnaseq.list_datasets()
datasets[["name", "title", "version"]].head(3)
```

:::

### Search for Datasets

You can also search for datasets based on metadata using `search_datasets()` in Python or `searchDatasets()` in R. This supports both simple text queries and complex boolean expressions.

::: {.panel-tabset}

## R
```{r}
pancreas_ds <- searchDatasets("pancreas")
head(pancreas_ds[, c("name", "title", "version")], 3)
```

## Python
```{python}
import scrnaseq

pancreas_datasets = scrnaseq.search_datasets("pancreas")
pancreas_datasets[["name", "title", "version"]].head(3)
```

:::

#### Advanced Searches

For more complex searches involving boolean operations, use `define_text_query()` in Python or `defineTextQuery()` in R. Here’s an example to find datasets using the mouse reference genome (`GRCm38`) and containing the words `neuro` or `pancrea`.

::: {.callout-tip}
The `define_text_query()` function in Python or its equivalent `defineTextQuery()` in R helps us define complex queries. Check out the reference manual for more details.
:::

::: {.panel-tabset}

## R
```{r}
suppressWarnings(library(gypsum))
res <- searchDatasets(
    defineTextQuery("GRCm38", field="genome") &
    (defineTextQuery("neuro%", partial=TRUE) | 
     defineTextQuery("pancrea%", partial=TRUE))
)
head(res[,c("name", "title", "version")], 3)
```

## Python
```{python}
from gypsum_client import define_text_query
import scrnaseq

res = scrnaseq.search_datasets(
     define_text_query("GRCm38", field="genome")
     & (
          define_text_query("neuro%", partial=True)
          | define_text_query("pancrea%", partial=True)
     )
)
res[["name", "title", "version"]].head(3)
```

:::

::: {.callout-important}
Once a dataset is identified, always list the name and version of the dataset in your scripts for reproducibility.
:::

## Fetch a Dataset

After identifying a dataset of interest, use `fetch_dataset()` in Python or `fetchDataset()` in R to download the dataset. This will load the dataset as a `SingleCellExperiment` object.

::: {.callout-note}
R/Bioconductor users might already be familiar with the [SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) class. BiocPy also provides the same implementation in the [singlecellexperiment](https://github.com/BiocPy/SingleCellExperiment) package.
:::

For this tutorial, let's download the `zeisel-brain` dataset:

::: {.panel-tabset}

## R
```{r}
sce <- fetchDataset("zeisel-brain-2015", "2023-12-14", realize.assays=TRUE)
sce
```

## Python
```{python}
import scrnaseq
sce = scrnaseq.fetch_dataset("zeisel-brain-2015", "2023-12-14", realize_assays=True)
print(sce)
```

:::

### Side-quest on SingleCellExperiment in Python

The Python version of the `SingleCellExperiment` class adheres to Bioconductor's specification and offers similar methods. Our goal is to make it simple for analysts to switch between R and Python. Key differences include a shift from functional to object-oriented paradigms.

::: {.callout-note}
For more details on the design, refer to the [BiocPy developer guide](https://github.com/BiocPy/developer_guide) or the [singlecellexperiment](https://github.com/BiocPy/SingleCellExperiment) documentation.
:::

```{python}
## repeated because quarto's build does not keep state of python snippets across the notebook.
import scrnaseq
sce = scrnaseq.fetch_dataset("zeisel-brain-2015", "2023-12-14")

print("assays: ", sce.get_assay_names()) # or sce.assay_names

print("column names: ", sce.get_column_names()) # or sce.column_names

print("column metadata:", sce.get_column_data()) # or sce.column_data

print("access counts ", sce.assays["counts"]) # or # sce.assay("counts")

print("coerce to AnnData", sce.to_anndata())
```

## Annotate Cell Types

We can now annotate cell types by using reference datasets and matching cells based on their expression profiles. In this tutorial, we will use [singleR](https://github.com/SingleR-inc/SingleR) in R or its Python equivalent [singler](https://github.com/BiocPy/singler).

Before running the singler algorithm, we need to fetch reference datasets from the `celldex` package.

## Access Reference Datasets from `celldex`

Similar to `scRNAseq`, the `celldex` references are converted into language-agnostic representations for use in downstream analyses.

::: {.callout-note}
The `celldex` package is available on [R/Bioconductor](https://bioconductor.org/packages/devel/data/experiment/html/celldex.html) and [PyPI](https://github.com/BiocPy/celldex).
:::

For this tutorial, let's download the Human Primary Cell Atlas reference from `celldex` using `fetch_reference()` in Python or `fetchReference()` in R.

::: {.panel-tabset}

## R
```{r}
suppressWarnings(library(celldex))
immgen_ref <- fetchReference("immgen", "2024-02-26", realize.assays=TRUE)
immgen_ref
```

## Python
```{python}
import celldex

immgen_ref = celldex.fetch_reference("immgen", "2024-02-26", realize_assays=True)
print(immgen_ref)
```

:::

Now, let's identify cells from the `zeisel-brain` dataset using the `immgen` reference dataset.

::: {.panel-tabset}

## R
```{r}
suppressWarnings(library(SingleR))
# sce <- scuttle::logNormCounts(sce)
cell_labels <- SingleR(test = assay(sce, "counts")[,1:100], ref = immgen_ref, labels = immgen_ref$label.main)

table(cell_labels$labels)
```

## Python
```{python}
import singler

import scrnaseq
sce = scrnaseq.fetch_dataset("zeisel-brain-2015", "2023-12-14", realize_assays=True)

import scranpy
options = scranpy.LogNormCountsOptions(with_size_factors=False)

# mat = sce.assays["counts"]
# log_normed_matrix = scranpy.log_norm_counts(mat_csc, options=options)

import celldex
immgen_ref = celldex.fetch_reference("immgen", "2024-02-26", realize_assays=True)

matches = singler.annotate_single(
    test_data=sce[:, :100], 
    ref_data = immgen_ref,
    ref_labels = "label.main"
)

import pandas as pd

pd.Series(matches["best"]).value_counts()
```

:::

Congratulations! You have now completed the tutorial on accessing single-cell datasets using `scRNAseq` and `ArtifactDB`, and annotating cell types using reference datasets from `celldex`. For more detailed usage and advanced analyses, refer to the respective documentation of these packages.
